Binning Strategies and Related Loss for Graphics of Large Data

### 2013-03-07 Paper.Rnw
Fixed figure problems I caused during the meeting yesterday.

      temp <- data.frame(   
                x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
             xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
                y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
             yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
            group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))
      qplot(data=d[,], x=x, y=y, alpha=a,  geom="point", 
            xlim=c(-20, 40), ylim=c(0, 50)) +
        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
                 fill="blue", alpha=.25) +
        annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), 
                 fill="green", alpha=.25) +
        scale_alpha(range=c(.5, 1), guide=FALSE) +
        scale_colour_identity(guide=FALSE) + 
        scale_shape(solid=T,guide=FALSE) +
        theme_bw() +
        annotate("point", x=rep(seq(-15, 35, 10),each=7), 
               y=rep(seq(-5, 55, 10), times=6), 
               shape='+', size=8, colour="gray90") + 
        geom_point(data=d[which(!is.na(d$xbin)),], 
                   aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), 
                   size=I(3), aes.inherit=FALSE) + 
        scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), 
                           labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
        scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", 
                            labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
        annotate("segment", x=temp$x, y=temp$y, 
                 xend=temp$xend, yend=temp$yend, group=temp$group)  + ylim(c(0, 50))
      #ggsave("./figure/Binning-DataAndBin.pdf", width=5, height=5, units="in")

      d.orig$bin <- dnew$x==15 & dnew$y==25
      d.orig$a <- .33+.66*d.orig$bin
      d.orig$l <- "Data outside bin"
      d.orig$l[d.orig$bin] <- "Binned Data"
      d.orig$size <- 2
      d.orig$size[d.orig$bin] <- 3
      d.orig <- rbind(d.orig, data.frame(x=15, y=25, a=1, l="Visual Center", size=2,bin=TRUE), 
                    data.frame(x=mean(d.orig$x[d.orig$bin]), y=mean(d.orig$y[d.orig$bin]), 
                               l="Numerical Center", a=1, size=2, bin=TRUE))
      ggplot(data=d.orig, aes(x=x, y=y, shape=l, size=l, colour=l)) + 
        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
                 fill="blue", alpha=.10) + 
        geom_point() +
        scale_shape_manual(name="", values=c("Data outside bin" = 16, 
                                             "Binned Data" = 15, 
                                             "Numerical Center" = 17, 
                                             "Visual Center" = 18)) +
        scale_colour_manual(name="", values=c("Data outside bin" = "grey60", 
                                              "Binned Data" = "blue",
                                              "Numerical Center" = "blue4", 
                                              "Visual Center" = "blue4"))+
        scale_size_manual(name="", values=c("Data outside bin" = 3, "Binned Data"=3,
                                            "Numerical Center" = 5, 
                                            "Visual Center" = 5))+
        theme_bw() +
        annotate("point", x=15, y=25, shape=18, size=5, colour="blue4") + 
        annotate("point", x=d.orig$x[which(d.orig$l=="Numerical Center")], 
                 y=d.orig$y[which(d.orig$l=="Numerical Center")], shape=17, size=5, 
                 colour="blue4") +
        annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + 
#         annotate("text", x=d$x[which(d$l=="Numerical Center")], 
#                  y=d$y[which(d$l=="Numerical Center")], 
#                  label=expression(paste('(', over(phantom(0),x)[i], ',', over(phantom(0),y)[i],')'))) +
        scale_x_continuous(breaks=c(10, 15, 20), limits=c(8, 22), labels=c(expression(x^{"*"}-b[x]/2), expression(x^{"*"}), expression(x^{"*"}+b[x]/2)), name='') + 
        scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
      #ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")

@@

\begin{figure}[hbtp]
  \centering
  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/binningDemo1.pdf}}
%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
%   {./figure/RandomBinningDataPoints1}}
  \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
  {./figure/RandomBinDemo.pdf}}
  \caption{Points are assigned to bin (0, 20) and bin (10,20). Points in the standard algorithm are all contained within the bin region, where points in the random algorithm may be outside the direct region.}
  \label{fig:BinningTutorial}
\end{figure}

@@

\begin{figure}[hbtp]
    \centering
		\subfloat[Spatial Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{binningDemo2}}
    \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.45\textwidth]{./figure/FreqLoss.pdf}}
\caption{\label{fig:stdBinTutorial2} Types of loss due to binning: (a) The visual center of the bin is different from the numerical center of the bin, leading to two different sources of loss.  (b) Frequency loss associated with a change of bin width to 1 from 0.25 in the example of Figure~\ref{binning}.}   
	\end{figure}

@@

\section{Discussion and Examples}
\subsection{Comparison of Binning Algorithms}
<<LossTableGraphs,echo=F,results="hide",cache=TRUE,fig.keep='none', eval=FALSE>>=
# connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
#                    password="R R0cks", port=3306, dbname="baseball", 
#                    host="headnode.stat.iastate.edu")
# pitch <- new("dataDB", co=connect, table="Pitching")
# 
# d1 <- dbData(pitch, vars=c( "G", "SO"))

@@

At the example of the pitching data, we will %consider the variables Games and Strikeouts, which are strongly correlated. This example 
highlight an important advantage of random binning. The two tile plots at the top of Figure~\ref{fig:binning-comparison} show bins of the same width, but clear differences between the binning results are evident: while the standard binning algorithm has rounding artifacts that are clearly visible for even bin widths, the random algorithm produces a graph that better represents the true shape of the data.
\begin{figure}[hbtp]
\centering
\subfloat[(2, 2) tiling of G, SO data]{
\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd210.pdf}
}\\
\subfloat[(5, 25) tiling of G, SO data]{
\includegraphics[keepaspectratio=true, width=.9\textwidth]{images/RandomStd525.pdf}
}
\caption{\label{fig:binning-comparison} Comparison of random and standard binning: standard binning introduces artificial striping when the chose bin width is not a multiple of the data resolution.}
\end{figure}

>>>>

      temp <- data.frame(   
                x=c(-20, -10,   0,  10,  20,  30,  40, -20, -20, -20, -20, -20, -20), 
             xend=c(-20, -10,   0,  10,  20,  30,  40,  40,  40,  40,  40,  40,  40), 
                y=c(  0,   0,   0,   0,   0,   0,   0,   0,  10,  20,  30,  40,  50), 
             yend=c( 50,  50,  50,  50,  50,  50,  50,   0,  10,  20,  30,  40,  50), 
            group=c(  1,   2,   3,   4,   5,   6,   7,   8,   9,  10,  11,  12,  13))
      qplot(data=d[,], x=x, y=y, alpha=a,  geom="point", 
            xlim=c(-20, 40), ylim=c(0, 50)) +
        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
                 fill="blue", alpha=.25) +
        annotate("polygon", x=c(0, 0, 10, 10), y=c(20, 30, 30, 20), 
                 fill="green", alpha=.25) +
        scale_alpha(range=c(.5, 1), guide=FALSE) +
        scale_colour_identity(guide=FALSE) + 
        scale_shape(solid=T,guide=FALSE) +
        theme_bw() +
        annotate("point", x=rep(seq(-15, 35, 10),each=7), 
               y=rep(seq(-5, 55, 10), times=6), 
               shape='+', size=8, colour="gray90") + 
        geom_point(data=d[which(!is.na(d$xbin)),], 
                   aes(x=x, y=y, shape=factor(xbin), colour=factor(xbin)), 
                   size=I(3), aes.inherit=FALSE) + 
        scale_shape_manual(name="Bin", values=c("5"=16, "15"=15), 
                           labels=c("(5,25)", "(15, 25)"),guide=FALSE) +
        scale_colour_manual(values=c("5"="green", "15"="blue"), name="Bin", 
                            labels=c("(5,25)", "(15, 25)"),guide=FALSE)+
        annotate("segment", x=temp$x, y=temp$y, 
                 xend=temp$xend, yend=temp$yend, group=temp$group)  + ylim(c(0, 50))
      ggsave("./figure/Binning-DataAndBin.pdf", width=5, height=5, units="in")

      d.orig$bin <- dnew$x==15 & dnew$y==25
      d.orig$a <- .33+.66*d.orig$bin
      d.orig$l <- "Data outside bin"
      d.orig$l[d.orig$bin] <- "Binned Data"
      d.orig$size <- 2
      d.orig$size[d.orig$bin] <- 3
      d.orig <- rbind(d.orig, data.frame(x=15, y=25, a=1, l="Visual Center", size=2,bin=TRUE), 
                    data.frame(x=mean(d.orig$x[d.orig$bin]), y=mean(d.orig$y[d.orig$bin]), 
                               l="Numerical Center", a=1, size=2, bin=TRUE))
      ggplot(data=d.orig, aes(x=x, y=y, shape=l, size=l, colour=l)) + 
        annotate("polygon", x=c(10, 10, 20, 20), y=c(20, 30, 30, 20), 
                 fill="blue", alpha=.10) + 
        geom_point() +
        scale_shape_manual(name="", values=c("Data outside bin" = 16, 
                                             "Binned Data" = 15, 
                                             "Numerical Center" = 17, 
                                             "Visual Center" = 18)) +
        scale_colour_manual(name="", values=c("Data outside bin" = "grey60", 
                                              "Binned Data" = "blue",
                                              "Numerical Center" = "blue4", 
                                              "Visual Center" = "blue4"))+
        scale_size_manual(name="", values=c("Data outside bin" = 3, "Binned Data"=3,
                                            "Numerical Center" = 5, 
                                            "Visual Center" = 5))+
        theme_bw() +
        annotate("point", x=15, y=25, shape=18, size=5, colour="blue4") + 
        annotate("point", x=d.orig$x[which(d.orig$l=="Numerical Center")], 
                 y=d.orig$y[which(d.orig$l=="Numerical Center")], shape=17, size=5, 
                 colour="blue4") +
        annotate("segment", x=temp$x, y=temp$y, xend=temp$xend, yend=temp$yend, group=temp$group) + 
#         annotate("text", x=d$x[which(d$l=="Numerical Center")], 
#                  y=d$y[which(d$l=="Numerical Center")], 
#                  label=expression(paste('(', over(phantom(0),x)[i], ',', over(phantom(0),y)[i],')'))) +
        scale_x_continuous(breaks=c(10, 15, 20), limits=c(8, 22), labels=c(expression(x^{"*"}-b[x]/2), expression(x^{"*"}), expression(x^{"*"}+b[x]/2)), name='') + 
        scale_y_continuous(breaks=c(20, 25, 30), limits=c(18, 32), labels=c(expression(y^{"*"}-b[y]/2), expression(y^{"*"}), expression(y^{"*"}+b[y]/2)), name='')
      ggsave("./figure/Binning-DataVisualNumericalCenter.pdf", width=5.5, height=3.25, units="in")

@@

\begin{figure}[hbtp]
  \centering
  \subfloat[Standard Algorithm]{\includegraphics[keepaspectratio=true, width=3in]{./figure/Binning-DataAndBin.pdf}}
%   \subfloat{\includegraphics[keepaspectratio=TRUE, width=2in]
%   {./figure/RandomBinningDataPoints1}}
  \subfloat[Random Algorithm]{\includegraphics[keepaspectratio=TRUE, width=3in]
  {./figure/RandomBinDemo.pdf}}
  \caption{Points are assigned to bin (0, 20) and bin (10,20). Points in the standard algorithm are all contained within the bin region, where points in the random algorithm may be outside the direct region.}
  \label{fig:BinningTutorial}
\end{figure}

@@

\begin{figure}[hbtp]
    \centering
		\subfloat[Spatial Loss]{\includegraphics[keepaspectratio=true, width=.54\textwidth]{./figure/Binning-DataVisualNumericalCenter}}
    \subfloat[Frequency Loss]{\includegraphics[keepaspectratio=true, width=.45\textwidth]{./figure/FreqLoss.pdf}}
\caption{\label{fig:stdBinTutorial2} Types of loss due to binning: (a) The visual center of the bin is different from the numerical center of the bin, leading to two different sources of loss.  (b) Frequency loss associated with a change of bin width to 1 from 0.25 in the example of Figure~\ref{binning}.}   
	\end{figure}

@@

\section{Discussion and Examples}
\subsection{Comparison of Binning Algorithms}
<<LossTableGraphs,echo=F,results="hide",cache=TRUE,fig.keep='none'>>=
# connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
#                    password="R R0cks", port=3306, dbname="baseball", 
#                    host="headnode.stat.iastate.edu")
# pitch <- new("dataDB", co=connect, table="Pitching")
# 
# d1 <- dbData(pitch, vars=c( "G", "SO"))

@@

At the example of the pitching data, we will %consider the variables Games and Strikeouts, which are strongly correlated. This example 
highlight an important advantage of random binning. The two tile plots at the top of Figure~\ref{fig:binning-comparison} show bins of the same width, but clear differences between the binning results are evident: while the standard binning algorithm has rounding artifacts that are clearly visible for even bin widths, the random algorithm produces a graph that better represents the true shape of the data.
\begin{figure}[hbtp]
\centering
\subfloat[(2, 2) tiling of G, SO data]{
\includegraphics[keepaspectratio=true, width=.9\textwidth]{figure/RandomStd210.pdf}
}\\
\subfloat[(5, 25) tiling of G, SO data]{
\includegraphics[keepaspectratio=true, width=.9\textwidth]{figure/RandomStd525.pdf}
}
\caption{\label{fig:binning-comparison} Comparison of random and standard binning: standard binning introduces artificial striping when the chose bin width is not a multiple of the data resolution.}
\end{figure}
